R Cheatsheet
============

Online References
-----------------

[Quick-R](http://www.statmethods.net)

[Cookbook for R](http://www.cookbook-r.com)

[An R Introduction to Statistics](http://www.r-tutor.com)

[R-seek](http://www.rseek.org)

Commonly Used Libraries
-----------------------

`lattice` - Lattice Graphics
`ggplot2` - Grammar of Graphics
`MASS` - Support datasets for Venables and Ripley text
`stats` - Common statistical functions

```{r}
library(lattice)
library(ggplot2)
library(MASS)
library(stats)
```

Constants
---------

```{r}
LETTERS
letters
month.abb
month.name
pi
```

Useful Functions
----------------

`aggregate` - compute summary statistics of data subsets
`apply` - apply functions over array margins
`apropos` - find objects by partial name
`droplevels` - drop unused levels from a factor or, more commonly, from factors in a data frame.
`c` - combine values into a vector or list
`head` - return the first part of an object
`identical` - test for exact identity
`nrow`, `ncol` - the number of rows or columns
`subset` - return subsets of vectors, matrices or data frames which meet conditions
`t` - transpose
`table` - make a contingency table
`tail` - return the last part of an object
`which` - Give the TRUE indices of a logical object, allowing for array indices.
`with` - Evaluate an R expression in an environment constructeed from data

Statistical Tests
-----------------

`t.test` - Performs one and two sample t-tests on vectors of data.

