R Cheatsheet
============

Online References
-----------------

[An R Introduction to Statistics](http://www.r-tutor.com)

Commonly Used Libraries
-----------------------

`lattice` - Lattice Graphics
`ggplot2` - Grammar of Graphics
`MASS` - Support datasets for Venables and Ripley text
`stats` - Common statistical functions

```{r}
library(lattice)
library(ggplot2)
library(MASS)
library(stats)
```

Constants
---------

```{r}
LETTERS
letters
month.abb
month.name
pi
```

Useful for generating lists like so:

```{r}
(names <- seq_along(letters[1:10]))
```


Useful Functions
----------------

`apropos` - Find objects by partial name

```{r}
apropos("^geom_")
```

`head`, `tail` - return the first or last part of an object

```{r}
head(letters)
tail(letters)
```

`identical` - test for exact identity

```{r}
identical(1, 1) # TRUE
identical(1, 2) # FALSE
```

`table` - make a contingency table

```{r}
table(rpois(100, 5))
```

`which` - Give the TRUE indices of a logical object, allowing for array indices.

```{r}
which(c(TRUE, FALSE))
```

`with` - Evaluate an R expression in an environment constructeed from data

```{r}
library(MASS)
with(anorexia, {
    anorex.1 <- glm(Postwt ~ Prewt + Treat + offset(Prewt),
                    family = gaussian)
    summary(anorex.1)
})
```

